---
title: "pivot_lobsters"
author: "Erin Aylsworth"
date: '2023-01-13'
output: html_document
---

```{r setup, include=FALSE}

library(tidyverse)
library(readxl)
library(here)
library(skimr) # install.packages("skimr")
library(kableExtra) # install.packages("kableExtra")

```


### Read in the first file: lobsters.xlsx

``` {r} 
lobsters <- read_xlsx(here("data/lobsters.xlsx"), skip = 4)
```

### 6.2.2.1 skimr::skim

Let's look at the summary statistics with skimr.

``` {r}

# explore data

skimr::skim(lobsters)

```

There are three character variables and four numeric.

### 6.2.3 Our task

We have been asked to report about how the average size of lobsters has changed for each site across time.

We will complete this task with R by using the dplyr packages for data wrangling, which we will do after dmeoing how this would do it with pivot tables in Excel.

### 6.4.1 group_by one variable

Let's use group_by() %>% summarize() with our lobsters data, just like we did in Excel.

``` {r}

lobsters |>
        group_by(year) |>
        summarize(count_by_year = n())

```

### 6.4.2 group_by multiple variables

``` {r}

lobsters |>
        group_by(site, year) |>
        summarize(count_by_siteyear = n())

```

### 6.4.2 summarize() multiple variables

We can summarize multiple variables at a time.

``` {r}

lobsters |>
        group_by(site, year) |>
        summarize(count_by_siteyear = n(),
                  mean_size_mm = mean(size_mm))
```

Oooooops, there are missing values, or NA. We can remove them and save the output for further investigation:

```{r}

siteyear_summary <- lobsters |>
        group_by(site, year) |>
        summarize(count_by_siteyear = n(),
                  mean_size_mm = mean(size_mm, na.rm = TRUE),
                  sd_size_mm = sd(size_mm, na.rm = TRUE))
```

Let's inspect our new variable:

```{r} 

## inspect our new variable

siteyear_summary

```

### 6.4.4 Table formatting with kable()

There are several options for formatting tables in RMArkdown; we'll show one here from the kableExtra package and learn more about it tomorrow.

It works nicely with the pipe operator, so we can build this from our new object:

``` {r}

## make a table with our new variable

siteyear_summary |>
        kable()

```

There are `r nrow(lobsters)` total lobsters included in this report.

### 6.4.6 Activity on your own!

Build from our analysis and calculate the mean lobster size for each site year.

Create and ggsave a plot

Then save, commit, and push your .Rmd, .html, and .png.

OK, here goes my attempt to first get the median

``` {r} 

my_summary <- lobsters |>
        group_by(site, year) |>
        summarize(count_by_siteyear = n(),
                  median_size_mm = median(size_mm, na.rm = TRUE))

ggplot(data = my_summary, aes(x = year, y = median_size_mm, color = site)) +
        geom_line()

ggsave(here("figures", "lobsters_median.png"))

```


